Enabling Ambient Backscatter \\Using a Low-Cost Software Defined Radio

### 2017-04-07 paper.tex
Aggressive reduction of design section

\subsection{Transmitter}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\columnwidth]{./fig/s11v2}
\caption{Input reflection coefficient \ensuremath{\Gamma_{in}} also known as \ensuremath{S_{\text{11}}} of the self-made backscattering ground-plane antenna.}
\label{fig:s11}
\end{figure}

Our backscattering antenna is optimized for usage with the television signal,
which has its center frequency at 626 MHz. It is a ground plane antenna, which
has been built up on a large unetched PCB plate. A hole in the middle of the
PCB serves for connecting a wire to a 50 \ensuremath{\Omega} micro strip
transmission line. The wire, which acts as a vertical antenna element is
soldered to the transmission line. The microstrip transmission line design
equations given in \cite{pozar_microwave_2011} are implemented in various
programs e.g.  KiCAD, which has been used for designing them. An open-stub,
attached directly at the juncture where the vertical antenna element is
attached to the transmission line, is used for fine tuning of the input
impedance (cf. chapter 5 in \cite{pozar_microwave_2011}).  Accepting -15 dB as
sufficient value for \ensuremath{S_{11}}, one can say, that the antenna works
within a range from 626 MHz to 645 MHz. With our narrowband antenna the
undesired backscattering of out-of-band signals is reduced. With our approach
we were able to achieve an input impedance at 634 MHz of
\ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz
\ensuremath{S_{\text{11}}} is still approximately -17 dB.

Besides the antenna, the transmitter consists of an \textit{MSP430} MCU,
that sends data to the receiver. This transmitter controls an RF switch
through an I/O pin. The switch can alternate the impedance connected to
the antenna between matched (50 \ensuremath{\Omega}) and open. The
software now steers this switch with a two 2 MHz rectangular timer
signal to transmit a data 1 or it just leaves the antenna open to
transmit a data 0. Therefore the result is a frequency shift of the
television signal by 2 MHz for a 1 and no shift for a 0. This can be
easily understood thinking about a simple cosine, which is shifted by
another cosine: \ensuremath{2\,cos(a)cos(b)=cos(a-b)+cos(a+b)}. Of
course this example is simplified compared to the actual situation as
the actual signal, which is used for shifting contains more than one
frequency as well as the TV signal, which is shifted. Hence a classical
binary amplitude shift keying is the implemented modulation technique.

\subsection{Receiver}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/receiver_arch}
\caption{Receiver architecture from a system point of view. Signal flow is from left to right.}
\label{fig:receiver_arch}
\end{figure}
We provide our C++ receiver code publicly available under
\url{https://github.com/s3xm3x/backscatterBASKReceiver}. A highly sophisticated
bus system has been developed to exchange data between the different components
of the system. The signal flow in figure \ref{fig:receiver_arch} is from the
left to the right.  \textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is
based on \textit{libusb}, is used to transfer the data from the TV stick into
our program. As described in section~\ref{sub:rtl2832}, the RTL2832 mixes down
the high frequency to a intermediate frequency (both values are software
adjustable). The sampled data is available as two 8-bit values (real and
imaginary). With this two values, the phase and absolute value of the data can
be determined for every sample. Every sample can be intepreted as:
\begin{equation}
	I+j\,Q = abs(I,Q) \cdot e^{j \times ang(I,Q)} 
\end{equation}

So the first block, which is entitled RTL2832U, provides the interface to the
TV dongle. This block controls the frequency \ensuremath{f_{tuned}}, where the
receiver is listening, besides other interesting parameters e.g. the analog
gain.

The demodulator block is responsible for converting the sampled signals into
ones and zeroes. Therefore the real and imaginary values first have to be
deinterleaved out of the sample message. These 8 bit integers are converted
into floats, whereas 127.5 equals 0. After that the values have to be
downsampled. Usually we were sampling with 250 kHz. This is still much more
than we need assuming an Additive White Gaussian Noise (AWGN) channel with a
\ensuremath{S/N = 10\,dB} (cf.  equation \ref{eq:awgn}).  So we decided to
reduce the sampling frequency to 25 kHz.

\begin{equation}
	\label{eq:awgn}
	C=B \cdot log_2(1+\frac{S}{N})= 25\,kHz \cdot log_2(11) \approx 86.5\,kbit/s
\end{equation}     
To do this it is important to use an anti-aliasing filter, as the Shannon-Nyquist theorem has to be satisfied: 
\begin{equation}
	\label{eq:nyquist}
	f_{samp} \geq 2 \cdot f_{a}
\end{equation}
\ensuremath{f_{a}} is the highest frequency in the signal. 
 A FIR filter
naively implemented with floats has been used to achieve this.  After
downsampling another filter is used to suppress noise.  The last demodulation
step is to rectify the signal (cf. equation \ref{eq:rectify}) and decide with a
software-defined Schmitt trigger whether a sample is a 1 or a 0.  
\begin{equation}
	\label{eq:rectify}
	abs(I,Q) = \sqrt{I^2 +  Q^2}
\end{equation} 
A registered listener receives the processed samples, which are broadcasted by
the demodulator. This registered listener is the decoder. After the decoder
received the samples it decides when a frame is sent or when the channel is
idle.  The decoder also includes a function to correlate the received bitstream
with an expected pattern. Hence it is able to determine the bit error ratio
(BER). Furthermore the code we have written so far subsumes different
simulators for e.g. playing back recoreded data. Another handy utility, which
has been written in Octave, is a tool (an oscilloscope) to look at the data,
which is written into files by the demodulator (cf. figure \ref{fig:transmission}).
\begin{figure}[h]

\centering
\includegraphics[width=\columnwidth]{./fig/transmission}
\caption{Start of a transmission of 101010.. with 1 kbit/s with idle state before. Data from the Octave oscilloscope. Amplitude is quite low with an average of 12 \% of the maximum, which is 127.5. This is due to a low signal strength. Sampling frequency is 25 kHz.}
\label{fig:transmission}
\end{figure}

@@

We have used the \textit{RTL-SDR} to observe the space variations of the
signal strength. \textit{Octave} scripts have been written to aid in
this purpose. These scripts are available online under
\url{https://github.com/s3xm3x/RTLSDRspecAn}. The scripts perform a
frequency sweep through the desired band. The obtained samples are then
transferred from the time domain to the frequency domain using a fast
fourier transformation (FFT).

>>>>

In this section we describe the most important aspects of our design.

\subsection{Transmitter}
%\begin{figure}[h]
%\centering
%\includegraphics[width=0.6\columnwidth]{./fig/s11v2}
%\caption{Input reflection coefficient \ensuremath{\Gamma_{in}} also known as \ensuremath{S_{\text{11}}} of the self-made backscattering ground-plane antenna.}
%\label{fig:s11}
%\end{figure}

%A backscattering antenna is the most important part of our backscatter
%transmitter. 
Our backscatter transmitter is designed to be used with the television
signal, which has its center frequency at 626 MHz. To that end, we build
an
antenna using an unetched PCB plate as the ground plane and a wire that acts
as the vertical antenna element.% is soldered to the transmission line. The
%transmission line has been calculated for 626 MHz using an FR4 substrate
%with a thickness of 1.6 mm. The design equations given in
%\cite{pozar_microwave_2011} are implemented in various programs e.g.
%KiCAD, which has been used for solving the design equations. An
%open-stub, attached directly at the juncture where the vertical antenna
%element is attached to the transmission line, is used for fine tuning of
%the input impedance (cf. chapter 5 in \cite{pozar_microwave_2011}).
%Accepting -15 dB as sufficient value for \ensuremath{S_{11}}, one can
%say, that the antenna works in a range from 626 MHz to 645 MHz. For
%backscattering this small bandwidth is not necessarily a disadvantage,
%since everything that is within the operating frequency of the antenna
%is backscattered. With our narrowband antenna the undesired
%backscattering of out-of-band signals is reduced.  With our approach we
%were able to achieve an input impedance at 634 MHz of
%\ensuremath{Z_{in}=(52.2 + j 4.3) \Omega}. At 626 MHz
%\ensuremath{S_{\text{11}}} is still approximately -17 dB.
This antenna works well in the range from 626 to 245 MHz.

Besides the antenna, the transmitter consists of an \textit{MSP430} MCU,
that sends data to the receiver. This transmitter controls an RF switch
through an I/O pin. The switch can alternate the impedance connected to
the antenna between matched (50 \ensuremath{\Omega}) and open. The
software now steers this switch with a two 2 MHz rectangular timer
signal to transmit a data 1 or it just leaves the antenna open to
transmit a data 0. Therefore the result is a frequency shift of the
television signal by 2 MHz for a 1 and no shift for a 0. Hence a classical
binary amplitude shift keying is the implemented modulation technique.

\subsection{Receiver}
\begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/receiver_arch}
\caption{Receiver architecture from a system point of view. Signal flow is from left to right.}
\label{fig:receiver_arch}
\end{figure}
C++ has been used to implement the receiver. We provide the code
publicly available under
\url{https://github.com/s3xm3x/backscatterBASKReceiver}. 
%A highly
%sophisticated bus system has been developed to exchange data between the
%different components of the system. 
Figure
\ref{fig:receiver_arch} shows the architecture of our receiver, the
signal flows from the left to the right.
%\textit{Librtlsdr} (cf. \cite{steve-m_librtlsdr}), which is based on
%\textit{libusb}, is used to transfer the data from the TV stick into our
%program. As described in section~\ref{sub:rtl2832}, the RTL2832 mixes
%down the high frequency to a intermediate frequency (both values are
%software adjustable). Subsequently the data is mixed with sine and
%cosine, low-pass filtered and finally transferred to the digital domain.
The raw I/Q data is available as two 8-bit values (real and imaginary)
from the RTL-SDR. With
these two values, the phase and magnitude of the data can be
determined for every sample. The received signal can be represented as:
\begin{equation}
	I+j\,Q = abs(I,Q) \cdot e^{j \times ang(I,Q)} 
\end{equation}

So the first block, entitled RTL2832U, provides the interface to the TV
dongle. It can set the frequency \ensuremath{f_{tuned}}, where the
receiver is listening, and the analog gain.

The demodulator block is responsible for converting the sampled signals
into ones and zeroes. 
%Therefore the real and imaginary values first have
%to be deinterleaved out of the sample message. Now they are represented
%as integer values between 0 and 255 (8 bit). So one converts them to
%floats, whereas 127.5 equals 0. After that the values have to be
%downsampled. Usually we were sampling with 250 kHz. This is still much
%more than we need assuming an Additive White Gaussian Noise (AWGN)
%channel with a \ensuremath{S/N = 10\,dB} (cf.  equation \ref{eq:awgn}).
%So we decided to reduce the sampling frequency to 25 kHz.
%\begin{equation}
%	\label{eq:awgn}
%	C=B \cdot log_2(1+\frac{S}{N})= 25\,kHz \cdot log_2(11) \approx 86.5\,kbit/s
%\end{equation}     
%To do this it is important to use an anti-aliasing filter, as the Shannon-Nyquist theorem has to be satisfied: 
%\begin{equation}
%	\label{eq:nyquist}
%	f_{samp} \geq 2 \cdot f_{a}
%\end{equation}
%\ensuremath{f_{a}} is the highest frequency in the signal. 
 A FIR
%filter naively implemented with floats has been used to achieve this. 10
%kHz is the cut-off frequency of the first filter. After downsampling
%another filter is used to suppress noise. This filter has a cut-off
%frequency of 3 kHz. 
As part of the demodulation process the signal is down-sampled and
low-pass-filtered two times. The last demodulation step is to rectify
the signal (cf. equation \ref{eq:rectify}) and decide with a
software-defined Schmitt trigger whether a sample is a 1 or a 0.  
\begin{equation}
	\label{eq:rectify}
	abs(I,Q) = \sqrt{I^2 +  Q^2}
\end{equation}

The resulting values are sent through the bus to the decoder. After the
decoder received the samples it decides when a frame is sent or when the
channel is idle.  
The decoder also includes a function to correlate the
received bitstream with an expected pattern. Hence it is able to
determine the bit error ratio (BER). 
Furthermore the code we have
written so far subsumes different simulators for e.g. playing back
recoreded data. Another handy utility, which has been written in Octave,
is a tool (an oscilloscope) to look at the data, which is written into
files by the demodulator. An example of a recorded data stream with 25
kS/s can be seen in figure
\ref{fig:transmission}.  \begin{figure}[h]
\centering
\includegraphics[width=\columnwidth]{./fig/transmission}
\caption{Start of a transmission of 101010.. with 1 kbit/s with idle state before. Data from the Octave oscilloscope. Amplitude is quite low with an average of 12 \% of the maximum, which is 127.5. This is due to a low signal strength.}
\label{fig:transmission}
\end{figure}

@@

We have used the textit{RTL-SDR} to observe the space variations of the
signal strength. \textit{Octave} scripts have been written to aid in
this purpose. These scripts are available online under
\url{https://github.com/s3xm3x/RTLSDRspecAn}. The scripts perform a
frequency sweep through the desired band. The obtained samples are then
transferred from the time domain to the frequency domain using a fast
fourier transformation (FFT).
